% English figure captions (measured-campaign scope in Section~\ref{sec:data-provenance}).

\newcommand{\CaptionFigVisualDisclaimer}{%
Campaign descriptive means (Table~\ref{tab:empirical-summary}); per-seed variance unavailable from committed CSVs (see Section~\ref{sec:data-provenance}). RPL\_MQoS and RPL\_AER reflect a single valid seed (seeds~1--3 artefact). AER-MQoS mean over seeds~2--4. Visual scaling applied for readability.%
}

\newcommand{\CaptionFigOne}{%
\textbf{Conceptual schematic---not simulation output.} Reference architecture of AER-MQoS: QoS plane, trust/energy context plane, MCS, and experimental OCP~8 objective function. Module boundaries match the Cooja firmware tree.%
}

\newcommand{\CaptionFigTwo}{%
\textbf{Conceptual schematic---not simulation output.} Illustrative RPL DODAG (not the 25-node Cooja layout): DIO (down), DAO (up), preferred parent path to the sink/LBR. Class-to-level mapping $C3\!\rightarrow\!1,\ldots,C0\!\rightarrow\!4$ is in Section~\ref{sec:arch}.%
}

\newcommand{\CaptionFigThree}{%
Context coefficient $\gamma$ (and derived weights $\alpha=\gamma$, $\beta=100-\gamma$ in hundredths) per traffic class $C0$--$C3$ for \texttt{AER-MQoS} from \texttt{context.csv} (measured Cooja campaign; Section~\ref{sec:data-provenance}). The two curves are mirror images since $\alpha+\beta=1$. This figure shares data with Table~\ref{tab:empirical-summary} and is provided as a complementary visual reference.%
}

\newcommand{\CaptionFigFour}{%
Overall and $C3$ PDR (\%) on the lossy UDGM channel ($\mathrm{success\_ratio}{=}0.85$): per-seed boxplots and descriptive means from \texttt{pdr.csv}. \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigFive}{%
Measured application-layer end-to-end latency (ms) on the lossy channel (TX/RX pairing; not the ETX-derived MCS surrogate). Descriptive means from \texttt{latency.csv}. \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigSix}{%
Mean inter-arrival gap (ms) per class $C0$--$C3$ from \texttt{jitter.csv} (not RFC~3393 IPDV; $C3$ ring-highlighted). Large gaps reflect sparse Poisson arrivals over long runs. \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigSeven}{%
Internal NRE (instrumentation) per firmware tag from \texttt{energy.csv} (\texttt{nre\_proxy\_pct}); not Energest duty-cycle or joules. \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigEight}{%
Control-plane export proxy: \texttt{ctrl\_export\_rate} from \texttt{ctrl.csv} (METRIC/context export rate per node per minute). Time axis covers the initial bootstrap window ($\approx$0--1\,min of 30\,min simulation). \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigNine}{%
Route-adaptation proxy: cumulative \texttt{ctx\_update\_cumulative} from \texttt{stab.csv} versus simulated time (METRIC/CTX proxy). Time axis covers the initial bootstrap window ($\approx$0--1\,min of 30\,min simulation). \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigTen}{%
Temporal PDR stratification over epochs \texttt{epoch\_first\_half} and \texttt{epoch\_second\_half} from \texttt{sec.csv} (four firmware tags; not an attack scenario). \CaptionFigVisualDisclaimer%
}

\newcommand{\CaptionFigEleven}{%
Logging sanity check: PDR vs.\ offered-load quartiles for \texttt{learning\_on} strata in \texttt{learn\_or\_load.csv} (same \texttt{AER-MQoS} binary; not a learning ablation). \CaptionFigVisualDisclaimer%
}
